What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2022-12-28 paper/main.tex
images

\begin{example}
    In \cref{fig:ex2}, one can reason about two timelines: the upper timeline starts with $p$ at the first time step, followed by entering the grey box with one step of $!p$ and one step of $p$. Then at the end of the grey box, we reenter the box, with the next time step being $!p$, and so on. The lower timeline starts with one step of $!p$ and one step of $p$ outside the grey box, and then enter the infinite run of $!p$ and $p$ repeating.
    \begin{figure}[!h]
        \centering
        \includegraphics[scale=0.4]{examples/ex2/ex2.gv.png}
        \caption{Example timeline for specification ``$p$ oscillates every time step''}
        \label{fig:ex2}
    \end{figure}
\end{example}

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualization works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automata % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth, trim=1cm 3cm 1cm 3cm, clip]{img/algorithm_outline.gv.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

@@

\begin{example}
    The formula $\Phi = \always (a \limplies \eventually (\neg a))$ generates un-simplified $\omega$-regular expression $((!a) | (aa^*(!a)))((!a) | (aa^*(!a)))^{\omega}$, and simplified version $((!a) | (aa^*(!a)))^{\omega}$, which corresponds to the two timelines in \cref{fig:unsimplified} and \cref{fig:simplified}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9-unsimplified.gv.png}
        \caption{Timeline for un-simplified $\omega$-regular expression $((!a) | (aa^*(!a)))((!a) | (aa^*(!a)))^{\omega}$}
        \label{fig:unsimplified}
    \end{figure}
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9.gv.png}
        \caption{Timeline for simplified $\omega$-regular expression $((!a) | (aa^*(!a)))^{\omega}$}
        \label{fig:simplified}
    \end{figure}
\end{example}

@@

\section{Theoretical Analyses}
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_dfa does its input to output behaviour
    For any DFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_dfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_dfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.75]{img/uvw_dfa.gv.pdf}
        \caption{$u$ to $w$ path in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be travelled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

@@

\begin{example}
    \cite{ZR14} presents a model verification specification says ``[i]f a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft'' which corresponds to \ltl\ formula
    \[
        \always (\texttt{tsafe.TSAFE\_command1} \land \texttt{controller.CTR\_control\_1} \limplies \nextt (\neg \texttt{controller.CTR\_control\_1}))
    \]
    For simplicity, we swap the concrete atomic proposition to $a, b, c$ and get
    \[
        \always (a \land b \limplies \nextt (\neg b))
    \]
    For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex14}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex14/ex14.gv.png}
        \caption{Timeline for $\always (a \land b \limplies \nextt (\neg b))$}
        \label{fig:ex14}
    \end{figure}
\end{example}

\begin{example}
    Spot~\cite{SPOT-online} presents command-line tool for random \ltl\ formula generator \texttt{randltl}~\cite{duret.13.atva}. The following \ltl\ formula is generated from this tool.
    \[
    p_2 \land (\eventually (\always p_0 \strongrelease  1) \weakrelease \nextt(\always p_1 \land ((p_0 \liff p_2) \stronguntil \eventually p_0)))
    \]
    This formula is reasonably complicated, and hard for human to reason about directly. For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex13}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex13/ex13.gv.png}
        \caption{Timeline for $p_2 \land (\eventually (\always p_0 \strongrelease  1) \weakrelease \nextt(\always p_1 \land ((p_0 \liff p_2) \stronguntil \eventually p_0)))$}
        \label{fig:ex13}
    \end{figure}
\end{example}
\begin{remark}
    These two examples show that our tool can generate reasonably clear diagrams for both real-world formula and randomly generated formula.
\end{remark}

>>>>

\begin{example}
    In \cref{fig:ex2}, one can reason about two timelines: the upper timeline starts with $p$ at the first time step, followed by entering the grey box with one step of $!p$ and one step of $p$. Then at the end of the grey box, we reenter the box, with the next time step being $!p$, and so on. The lower timeline starts with one step of $!p$ and one step of $p$ outside the grey box, and then enter the infinite run of $!p$ and $p$ repeating.
    \begin{figure}[!h]
        \centering
        \includegraphics[scale=0.4]{examples/ex2/ex2.png}
        \caption{Example timeline for specification ``$p$ oscillates every time step''}
        \label{fig:ex2}
    \end{figure}
\end{example}

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualization works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automata % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth, trim=1cm 3cm 1cm 3cm, clip]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

@@

\begin{example}
    The formula $\Phi = \always (a \limplies \eventually (\neg a))$ generates un-simplified $\omega$-regular expression $((!a) | (aa^*(!a)))((!a) | (aa^*(!a)))^{\omega}$, and simplified version $((!a) | (aa^*(!a)))^{\omega}$, which corresponds to the two timelines in \cref{fig:unsimplified} and \cref{fig:simplified}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9-unsimplified.png}
        \caption{Timeline for un-simplified $\omega$-regular expression $((!a) | (aa^*(!a)))((!a) | (aa^*(!a)))^{\omega}$}
        \label{fig:unsimplified}
    \end{figure}
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9.png}
        \caption{Timeline for simplified $\omega$-regular expression $((!a) | (aa^*(!a)))^{\omega}$}
        \label{fig:simplified}
    \end{figure}
\end{example}

@@

\section{Theoretical Analyses}
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_dfa does its input to output behaviour
    For any DFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_dfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_dfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{$u$ to $w$ path in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be travelled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

@@

\begin{example}
    \cite{ZR14} presents a model verification specification says ``[i]f a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft'' which corresponds to \ltl\ formula
    \[
        \always (\texttt{tsafe.TSAFE\_command1} \land \texttt{controller.CTR\_control\_1} \limplies \nextt (\neg \texttt{controller.CTR\_control\_1}))
    \]
    For simplicity, we swap the concrete atomic proposition to $a, b, c$ and get
    \[
        \always (a \land b \limplies \nextt (\neg b))
    \]
    For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex14}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex14/ex14.png}
        \caption{Timeline for $\always (a \land b \limplies \nextt (\neg b))$}
        \label{fig:ex14}
    \end{figure}
\end{example}

\begin{example}
    Spot~\cite{SPOT-online} presents command-line tool for random \ltl\ formula generator \texttt{randltl}~\cite{duret.13.atva}. The following \ltl\ formula is generated from this tool.
    \[
    p_2 \land (\eventually (\always p_0 \strongrelease  1) \weakrelease \nextt(\always p_1 \land ((p_0 \liff p_2) \stronguntil \eventually p_0)))
    \]
    This formula is reasonably complicated, and hard for human to reason about directly. For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex13}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex13/ex13.png}
        \caption{Timeline for $p_2 \land (\eventually (\always p_0 \strongrelease  1) \weakrelease \nextt(\always p_1 \land ((p_0 \liff p_2) \stronguntil \eventually p_0)))$}
        \label{fig:ex13}
    \end{figure}
\end{example}
\begin{remark}
    These two examples show that our tool can generate reasonably clear diagrams for both real-world formula and randomly generated formula.
\end{remark}
